## brms

```{r}
#| echo: false
source("_params.r")
gryphon <- read.csv("data/gryphon.csv")
gryphon$animal <- as.factor(gryphon$animal)
gryphon$mother <- as.factor(gryphon$mother)
gryphon$byear <- as.factor(gryphon$byear)
gryphon$sex <- as.factor(gryphon$sex)
gryphon$bwt <- as.numeric(gryphon$bwt)
gryphon$tarsus <- as.numeric(gryphon$tarsus)
gryphonped <- read.csv("data/gryphonped.csv")
gryphonped$id <- as.factor(gryphonped$id)
gryphonped$father <- as.factor(gryphonped$father)
gryphonped$mother <- as.factor(gryphonped$mother)
```

### Running the model	
First, we need to load the `brms` library:

```{r}
library(brms)
```
To be able to fit an animal model, `brms` needs the relativeness (relationship) matrix of the pedigree and not its inverse (as in other softwares).
This can be estimated using the `nadiv` package created by Pr. Matthew Wolak (https://cran.r-project.org/web/packages/nadiv/index.html).

```{r}
Amat <- as.matrix(nadiv::makeA(gryphonped))
```

We are now ready to specify our first model:
The structure of a `bmrs` model is similar to `lme4`, thus the random effect is added to the model with the term `(1 | gr(animal, cov = Amat)` which associated the animal ID to the matrix of relativeness. 
In addition to the synthase of `lme4`, we include other features or parameters within the models such as `chain` which represent the number of Markov chains (defaults to 4), `core` which represents the number of cores to use when executing the chains in parallel and `iter` which represents the number of total iterations per chain. For more parameters such as `thin` or `warmup/burnin`, you can read the Cran R page of the package (https://cran.r-project.org/web/packages/brms/brms.pdf)

Given that `bmrs` fit the model using a Bayesian approach via the software `stan`, we need to specify priors for the model. Default priors in `brms` work relatively well, however we strongly suggest to carefully select an adequate prior for your analysis.
In this tutorial we will use the default priors. 
To get the prior used by default, we can use the `get_prior()` function.

```{r}
#| eval: !expr params$fit_all
brms_m1.1 <- brm(
  bwt ~ 1 + (1 | gr(animal, cov = Amat)),
  data = gryphon,
  data2 = list(Amat = Amat),
  family = gaussian(),
  chains = 1, cores = 1, iter = 100
)
#
save(brms_m1.1, file = "data/brms_m1_1.rda")
```

The result of the long model calculation is saved in a spare file `brms_m1_1.rda"`.
To help readers, we can directly  reload it.

```{r}
load("data/brms_m1_1.rda")
```

Two distinct plots can be produced to visualize the model diagnostic `mcmc_plot`. Note, that `sigma` represents the residual standard deviation.

Next, we examine (or directly using the model) the variance estimates and their distributions (via `summary` or plot). 

```{r}
plot(brms_m1.1)
mcmc_plot(brms_m1.1, type = "acf")
summary(brms_m1.1)
```

The `plot` of variance showed that the different variances have a normal distribution, the autocorrelation plot or 'acf' show that the autocorrelation is close to 0.
The `summary` exposes the  mean (Estimate) of each variance or fixed effect (here just the intercept) associated to their posterior distribution with standard deviation (Est.Error) and two-sided 95% Credible intervals. `Rhat` provides information on the estimate convergence. If it's greater than 1, the chains have not yet converged, and it will be required to run more iterations and/or set stronger priors. `ESS` represents the Effective sample values as the number of independent samples from the posterior distribution. However, for the purpose of this guide, the Rhat values are acceptable.
It is also possible to calculate the heritability using the function 'as.mcmc'.

```{r}
v_animal <- (VarCorr(brms_m1.1, summary = FALSE)$animal$sd)^2
v_r <- (VarCorr(brms_m1.1, summary = FALSE)$residual$sd)^2
h.bwt.1 <- as.mcmc(v_animal / (v_animal + v_r))
summary(h.bwt.1)
plot(h.bwt.1)
# or
Var.table <- as_draws_df(brms_m1.1)
Var.table$h.bwt.1 <- as.mcmc((Var.table$sd_animal__Intercept)^2 / ((Var.table$sd_animal__Intercept)^2 + (Var.table$sigma)^2))
summary(Var.table$h.bwt.1)
plot(Var.table$h.bwt.1)
```

### Adding fixed effects

To add effects to a univariate model, we simply modify the priors and the fixed effect portion of the model specification:

```{r}
#| eval: !expr params$fit_all
brms_m1.2 <- brm(
  bwt ~ 1 + sex + (1 | gr(animal, cov = Amat)),
  data = gryphon,
  data2 = list(Amat = Amat),
  family = gaussian(),
  chains = 2, cores = 2, iter = 1000
)

save(brms_m1.2, file = "data/brms_m1_2.rda")
```

To save time, result of the calculation is stored in the spare file `brms_m1_2.rda"`. 
We can assess the effect of `sex` as a fixed effect by examining its posterior distribution.

```{r}
load("data/brms_m1_2.rda")
summary(brms_m1.2)
plot(brms_m1.2)
mcmc_plot(brms_m1.2, type = "pairs")

summary(brms_m1.2)$fixed
```

The posterior distribution of the `sex2` term does not overlap zero. Thus, we can infer that sex has an effect on birth weight (presence of a sexual dimorphism) in this model and is a useful addition to the model, for most purposes. It is also worth noting that the variance components have changed slightly:
```{r}
summary(brms_m1.2)$random
```

In fact, since sex effects were previously contributing to the residual variance of the model our estimate of $V_R$ (denoted ’units’ in the output) is now slightly lower than before. This has an important consequence for estimating heritability since if we calculate $V_P$ as $V_A +V_R$ then as we include fixed effects, we will soak up more residual variance driving $V_P$ . Assuming that $V_A$ is more or less unaffected by the fixed effects fitted then as $V_P$ goes down we expect our estimate of $h^2$ will go up.
```{r}
v_animal <- (VarCorr(brms_m1.2, summary = FALSE)$animal$sd)^2
v_r <- (VarCorr(brms_m1.2, summary = FALSE)$residual$sd)^2
h.bwt.2 <- as.mcmc(v_animal / (v_animal + v_r))

summary(h.bwt.2)
summary(h.bwt.1)
```

Here $h^2$ has increased slightly from `r round(mean(h.bwt.1),2)` to `r round(mean(h.bwt.2),2)` (again, your values may differ slightly due to Monte Carlo error). Which is the better estimate? 
It depends on what your question is. The first is an estimate of the proportion of variance in birth weight explained by additive effects, the latter is an estimate of the proportion of variance in birth weight after conditioning on sex that is explained by additive effects.
An important piece of advice, each researcher should be consistent in how they name their estimates and always correctly describe which estimates they are using conditional or not (to avoid any confusion).


### Adding random effects

This is done by simply modifying the model statement, but it requires adding  a prior for the new random effect. For instance, we can fit an effect of birth year and for mother:

```{r}
#| eval: !expr params$fit_all
brms_m1.3 <- brm(
  bwt ~ 1 + sex + (1 | gr(animal, cov = Amat)) + (1 | byear),
  data = gryphon,
  data2 = list(Amat = Amat),
  family = gaussian(),
  chains = 2, cores = 2, iter = 1000
)

save(brms_m1.3, file = "data/brms_m1_3.rda")
```

To save time, the result of the calculation is stored in the spare file `brms_m1_3.rda"`.

```{r}
load("data/brms_m1_3.rda")
plot(brms_m1.3)
```

Here the variance in birth weight explained by birth year is `r round(VarCorr(brms_m1.3)$byear$sd^2,2)`.
Note that although $V_A$ has changed somewhat, most of what is now partitioned as a ‘birth year’ effect was previously partitioned as $V_R$.
Thus, what we have really done here is to partition environmental effects into those arising from year-to-year differences versus everything else, and we do not really expect much change in $h^2$ (since now $h^2 = V_A /(V_A + V_{BY} + V_R )$).
However, we get a somewhat different result if we also add a random effect of mother to test for maternal effects: 

```{r}
#| eval: !expr params$fit_all
brms_m1.4 <- brm(
  bwt ~ 1 + sex + (1 | gr(animal, cov = Amat)) + (1 | byear) + (1 | mother),
  data = gryphon,
  data2 = list(Amat = Amat),
  family = gaussian(),
  chains = 2, cores = 2, iter = 1000
)

save(brms_m1.4, file = "data/brms_m1_4.rda")
```

To save time, the result of the calculation is stored in the spare file `brms_m1_4.rda"`. 

```{r}
load("data/brms_m1_4.rda")
plot(brms_m1.4)
```

Here partitioning of significant birth year and maternal variance has resulted in a further decrease in $V_R$ but also a decrease in $V_A$.
The latter is because maternal effects of the sort we simulated (fixed differences between mothers) will have the consequence of increasing similarity among maternal siblings. Consequently, they can look very much like an additive genetic effect and if present, but unmodelled, represent a type of ‘common environment effect’ that can - and will- cause upward bias in $V_A$ and so $h^2$. 

```{r}
v_animal <- (VarCorr(brms_m1.4, summary = FALSE)$animal$sd)^2
v_byear <- (VarCorr(brms_m1.4, summary = FALSE)$byear$sd)^2
v_mother <- (VarCorr(brms_m1.4, summary = FALSE)$mother$sd)^2
v_r <- (VarCorr(brms_m1.4, summary = FALSE)$residual$sd)^2
h.bwt.4 <- as.mcmc(v_animal / (v_animal + v_byear + v_mother + v_r))
# or
Var.table <- as_draws_df(brms_m1.4)
Var.table$h.bwt.4 <- as.mcmc(
  Var.table$sd_animal__Intercept^2 /
    (
      Var.table$sd_animal__Intercept^2 + Var.table$sd_byear__Intercept^2 +
        Var.table$sd_mother__Intercept^2 + Var.table$sigma^2
    )
)
summary(Var.table$h.bwt.4)
plot(Var.table$h.bwt.4)
```

Let’s compare the estimates of heritability from each of models 1.1, 1.2 and 1.4:
```{r}
summary(h.bwt.4)
summary(h.bwt.2)
summary(h.bwt.1)
```

### Testing significance of variance components
While testing the significance of fixed effects by evaluating whether or not their posterior distributions overlap zero was simple and valid, this approach does not work for variance components. 
Variance components are bounded to be positive (given a proper prior), and thus even when a random effect is not meaningful, its posterior distribution will never overlap zero. 
Model comparisons can be performed using the function `loo_compare` using `waic` or weighted AIC.
```{r}
brms_m1.3 <- add_criterion(brms_m1.3, "loo")
brms_m1.1 <- add_criterion(brms_m1.1, "loo")

loo_compare(brms_m1.3, brms_m1.1, criterion = "loo")
```
Need more explication on the function ??

### Further partitioning of the variance
A population can be further fragmented into different groups or categories (such as females and males, juveniles and adults or treated and untreated). Some scientific questions require further and deeper analysis of the variance. To avoid multiple model (one for each group), we can directly partition the variance between groups in a unique model. In addition, by doing so, we can also test if the variance is different between groups.
As example, we can partition the additive genetic variance and residual variance by sex. It is impossible to further partition the other variances but complexity an animal model requires sufficient sample size.
Two distinct approaches can be done to partition the different random effect: using an extra argument `by=sex` or by adding `(0+sex|)` before the `|`. Notes, here we used `||` which not estimate a possible covariance between groups (female and male) for the random effect.

```{r}
#| eval: !expr params$fit_all
brms_m1.sex <- brm(
  bwt ~ 1 + sex +
    (0 + sex || gr(animal, cov = Amat)) +
    (0 + sex || byear) +
    (0 + sex || mother),
  data = gryphon,
  data2 = list(Amat = Amat),
  family = gaussian(),
  chains = 2, cores = 2, iter = 1000
)
save(brms_m1.sex, file = "data/brms_m1_sex.rda")
```

To save time, the result of the calculation is stored in the spare file `brms_m1_sex.rda"`. 
```{r}
load("data/brms_m1_sex.rda")
summary(brms_m1.sex)
```

We can see the model estimate variance for both sexes.  However, the residual level or sigma is not split by sexes. A further and more complex code need to be performed  to estimate the sex-specific heritability.
Note here, we first produce the model parameter in `bf_m2.sex` than implement it into `brms` command ( to show some other way to code in brms). 
```{r}
#| eval: !expr params$fit_all
bf_m1.sex2 <- bf(
  bwt ~ 1 + sex +
    (0 + sex || gr(animal, cov = Amat)) +
    (0 + sex || mother) +
    (0 + sex || byear),
  sigma ~ sex - 1
)

brms_m1.sex2 <- brm(bf_m1.sex2,
  data = gryphon,
  data2 = list(Amat = Amat),
  family = gaussian(),
  chains = 2, cores = 2, iter = 1000
)

save(brms_m1.sex2, file = "data/brms_m1_sex2.rda")
```

To save time, the result of the calculation is stored in the spare file `brms_m1_sex.rda"`. 
```{r}
load("data/brms_m1_sex2.rda")
summary(brms_m1.sex2)

Var.table <- as_draws_df(brms_m1.sex2)
Var.table$h.bwt.f <- as.mcmc(
  Var.table$sd_animal__sex1^2 /
    (
      Var.table$sd_animal__sex1^2 + Var.table$ sd_byear__sex1^2 +
        Var.table$sd_mother__sex1^2 + Var.table$b_sigma_sex1^2
    )
)
Var.table$h.bwt.m <- as.mcmc(
  Var.table$sd_animal__sex2^2 /
    (
      Var.table$sd_animal__sex2^2 + Var.table$ sd_byear__sex2^2 +
        Var.table$sd_mother__sex2^2 + Var.table$b_sigma_sex2^2
    )
)

summary(Var.table$h.bwt.f)
summary(Var.table$h.bwt.m)
plot(Var.table$h.bwt.f)
plot(Var.table$h.bwt.m)
```

Here, we can plot the point estimates of the $h^2$ which seems to differ between sexes, but their CI overlaps.
```{r}
#| fig-cap: Female and male heritability of birth weight
h2.sex <- rbind(
  cbind(
    summary(Var.table$h.bwt.f)$statistics[1],
    summary(Var.table$h.bwt.f)$quantiles[1],
    summary(Var.table$h.bwt.f)$quantiles[5]
  ),
  cbind(
    summary(Var.table$h.bwt.m)$statistics[1],
    summary(Var.table$h.bwt.m)$quantiles[1],
    summary(Var.table$h.bwt.m)$quantiles[5]
  )
)

plot(c(0.95, 1.05) ~ h2.sex[, 1],
  xlim = c(0, 0.8), ylim = c(0.5, 1.5),
  xlab = "", ylab = "", col = c("red", "blue"),
  pch = c(16, 17), cex = 2, yaxt = "n"
)
arrows(
  y0 = 0.95, x0 = h2.sex[1, 2],
  y1 = 0.95, x1 = h2.sex[1, 3],
  code = 3, angle = 90, length = 0, col = c("red"), lwd = 2
)
arrows(
  y0 = 1.05, x0 = h2.sex[2, 2],
  y1 = 1.05, x1 = h2.sex[2, 3],
  code = 3, angle = 90, length = 0, col = c("blue"), lwd = 2
)
mtext(
  "Narrow-sense heritability (±CI)",
  side = 1, las = 1, adj = 0.4, line = 3, cex = 1.6
)
axis(2, at = 1, labels = c("birth weight"), las = 3, cex.axis = 1.6)
```

### Modification of model parameter
Unfortunately (to our knowledge), it is not possible to alter the variance matrices and refit them within the model. 


### Covariance between two random effects
Some research questions require to estimate the covariance between two random effects within a univariate model.
Unfortunately (to our knowledge), it is not possible to create a covariance between distinct random effects (https://github.com/paul-buerkner/brms/issues/502).
However, a multi-membership model can be fit using the linking.function `mm`, thus forcing the variance of two variables to be equal and the covariance to 1. 


### Reaction norm and plasticity
Since we know how to partition the variance across different levels or group within population (sex as example). It is also possible to partition across environment (character stat approach) which allow to estimate the variance and covariance across levels of environment. There is an alternative way to visualize and estimate the change in variance between environment using a reaction norm known as plasticity. Instead to estimate variance at each level, we can modelized the slope of variance change across environments as a continuous variable.  
Note: Reaction norms are easier to estimate than character state approach, due to the fact that the reaction norm involved a univariate model with a slope whereas the character states approach involves a hybrid of multivariate model.  
For this example, we decide to use sex as an environment, thus estimating the reaction norm across sex, but usually researchers focus on environment.

```{r}
#| eval: !expr params$fit_all

bf_m1.RN <- bf(
  bwt ~ 1 + sex +
    (1 + sex | gr(animal, cov = Amat)) +
    (1 | mother) + (1 | byear)
)

brms_m1.RN <- brm(bf_m1.RN,
  data = gryphon,
  data2 = list(Amat = Amat),
  family = gaussian(),
  chains = 2, cores = 2, iter = 1000
)

save(brms_m1.RN, file = "data/brms_m1_RN.rda")
```

To save time, the result of the calculation is stored in the spare file `brms_m1_RN.rda"`. 
```{r}
load("data/brms_m1_RN.rda")

summary(brms_m1.RN)
plot(brms_m1.RN)
summary(brms_m1.RN)$fixed
summary(brms_m1.RN)$random
```
To implement a reaction norm between groups, we use the function `(1 + sex | gr(animal, cov = Amat):` which estimate the intercept, slope of the reaction norm, and interaction between both.  Here the parameter names are easy to understand (intercept, slope and correlation between both).

Interestingly, both models ( reaction norm and character states) are intertown in their equation, and a possible back transformation can be executed. `Part in progress, we try the hard work`

```{r}
RN_Matrix <- matrix(NA, ncol = 2, nrow = 2)
RN_Matrix[upper.tri(RN_Matrix, diag = T) == T] <- as.numeric(
  summary(brms_m1.RN)$random$animal[1:3, 1]
)
RN_Matrix_2 <- t(RN_Matrix)
RN_Matrix[lower.tri(RN_Matrix, diag = F) == T] <- RN_Matrix_2[lower.tri(
  RN_Matrix_2,
  diag = F
) == T]
RN_Matrix
Q <- matrix(c(1, 1, -1, 1), nrow = 2, ncol = 2)
tQQinv <- solve(crossprod(Q))
CS_Matrix <- tQQinv %*% t(Q) %*% RN_Matrix %*% Q %*% t(tQQinv)
```
